Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery

### 2025-03-06 paper/markdown/manuscript.Rmd
Variable redundancy
Inclusion of variable redundancy analysis data in Results of manuscript. modification of Supplementary Figure S14

Reduced DLCO, FVC and FEV1 (each < 80% of reference), as well as values of DLCO, FVC and FEV1 expressed as percentages of the reference value were modeled with `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` explanatory variables. 
The explanatory variables included 
demographic features (e.g. age, sex, body mass index, smoking, comorbidity), 
characteristic of acute COVID-19 (severity, medication) and recovery (e.g. weight loss, symptoms of relevance for lung function, time after diagnosis), 
and presence of human- and AI-rated structural lung abnormalities in CT scans (e.g. GGO, CTSS, opacity and high opacity). 
The modeling responses and explanatory variables are listed in __Supplementary Table S2__. 
The models were constructed with four common machine learning algorithms: 
Random Forest [@Wright2017; @Breiman2001], 
gradient boosted machines (GBM) [@Greenwell2022; @Friedman2002; @Friedman2001], 
neural network with a single hidden layer [@Ripley2014], 
and support vector machines (SVM) with radial kernel [@Weston1998; @Karatzoglou2004]. 
Selection of the optimal values of parameters controlling model behavior such as number of random trees, neurons in the hidden layer, or cost parameter was motivated by the maximum of Youden's J statistic (classification models of LFT abnormalities) or minimum mean absolute error (MAE, regression models of LFT readouts) in 10-repeats 10-fold cross-validation [@Kuhn2008]. 
Because of the presence of participant-matched observation, blocked cross-validation design was used both in the model selection and model evaluation, with blocks defined by participant's identifier. 
Model predictions were evaluated both in the training data and blocked 10-repeats 10-fold cross-validation. 
Concordance between the model-predicted and observed outcomes for classification models was assessed by Cohen's $\kappa$ inter-rater reliability statistic [@McHugh2012; @Cohen1960]. 
Accuracy, AUC, specificity and sensitivity of the classification model were investigated by ROC. 
Calibration of the classification models was assessed by Brier scores [@Brier1950]. 
Fraction of explained variance in predictions of the regression models was measured by pseudo-R^2^, the regression model error was expressed as MAE. 
Spearman's $\rho$ coefficients of correlation between the predicted and observed values were used to gauge calibration of the regression models. 
Over- and under-fitting was assessed by learning curves [@Viering2023]. 
Importance of explanatory variables was estimated by absolute values of SHAP statistics (Shapley additive explanations) [@Covert2020; @Lundberg2017; @Mayer2023a]. 
Co-linearity of the most influential explanatory variables was assessed by weighted correlation graph, whose edge weights corresponded to Kendall's $\tau$ coefficients of pairwise correlations [@Csardi2006].

@@

To assess under- and over-fitting, we resorted to analyses of learning curves of the models of DLCO and reduced DLCO. With this approach, we evaluated performance of the models re-trained in subsets of the modeling data set of varying sizes. 
The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
The learning curves for the best-performing GBM models are presented in __Supplementary Figure S12__, for the remaining algorithms comparable results were obtained (not shown). 
As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e. poor generalizability for unseen data. 
The generalizability of the models of DLCO expressed as percentage of the patient#s reference was substantially better.

@@

***placeholder for colinearity of the explanatory varibales***

>>>>

Reduced DLCO, FVC and FEV1 (each < 80% of reference), as well as values of DLCO, FVC and FEV1 expressed as percentages of the reference value were modeled with `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` explanatory variables. 
The explanatory variables included 
demographic features (e.g. age, sex, body mass index, smoking, comorbidity), 
characteristic of acute COVID-19 (severity, medication) and recovery (e.g. weight loss, symptoms of relevance for lung function, time after diagnosis), 
and presence of human- and AI-rated structural lung abnormalities in CT scans (e.g. GGO, CTSS, opacity and high opacity). 
The modeling responses and explanatory variables are listed in __Supplementary Table S2__. 
The models were constructed with four common machine learning algorithms: 
Random Forest [@Wright2017; @Breiman2001], 
gradient boosted machines (GBM) [@Greenwell2022; @Friedman2002; @Friedman2001], 
neural network with a single hidden layer [@Ripley2014], 
and support vector machines (SVM) with radial kernel [@Weston1998; @Karatzoglou2004]. 
Selection of the optimal values of parameters controlling model behavior such as number of random trees, neurons in the hidden layer, or cost parameter was motivated by the maximum of Youden's J statistic (classification models of LFT abnormalities) or minimum mean absolute error (MAE, regression models of LFT readouts) in 10-repeats 10-fold cross-validation [@Kuhn2008]. 
Because of the presence of participant-matched observation, blocked cross-validation design was used both in the model selection and model evaluation, with blocks defined by participant's identifier. 
Model predictions were evaluated both in the training data and blocked 10-repeats 10-fold cross-validation. 
Concordance between the model-predicted and observed outcomes for classification models was assessed by Cohen's $\kappa$ inter-rater reliability statistic [@McHugh2012; @Cohen1960]. 
Accuracy, AUC, specificity and sensitivity of the classification model were investigated by ROC. 
Calibration of the classification models was assessed by Brier scores [@Brier1950]. 
Fraction of explained variance in predictions of the regression models was measured by pseudo-R^2^, the regression model error was expressed as MAE. 
Spearman's $\rho$ coefficients of correlation between the predicted and observed values were used to gauge calibration of the regression models. 
Over- and under-fitting was assessed by learning curves [@Viering2023]. 
Importance of explanatory variables was estimated by absolute values of SHAP statistics (Shapley additive explanations) [@Covert2020; @Lundberg2017; @Mayer2023a]. 
Co-linearity of the most influential explanatory variables (top 15 mean absolute SHAP for each of the models of DLCO and DLCO < 80%) was assessed by soft-threshold weighted graph of correlations [@Csardi2006; @Zhang2005]. 
The graph edges were defined by pairwise correlations with Kendall's $\tau \geq$ 0.3 and edge weights corresponded to $\tau$ coefficient values.

@@

To assess under- and over-fitting, we resorted to analyses of learning curves of the models of DLCO and reduced DLCO. With this approach, we evaluated performance of the models re-trained in subsets of the modeling data set of varying sizes. 
The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
The learning curves for the best-performing GBM models are presented in __Supplementary Figure S12__, for the remaining algorithms comparable results were obtained (not shown). 
As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e. poor generalizability for unseen data. 
The generalizability of the models of DLCO expressed as percentage of the patient's reference was substantially better.

@@

Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S14A__). 
In particular, CTSS and opacity (correlation coefficient, Spearman's $\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'opacity_percent') %>% .$rho %>% signif(2)`), CTSS and high opacity ($\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`), and opacity and high opacity (`r corr_lft$test$ct %>% filter(variable1 == 'opacity_percent', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`) were strongly inter-correlated. 
CTSS values were significantly higher in observations with than without 
GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$eff_size`) and 
reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$eff_size`) (__Supplementary Figure S3B__, __Supplementary Table S10__). 
Analogically, significantly higher opacity was detected in observations with than without 
GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$eff_size`) and 
reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$eff_size`). 
Effect sizes of these differences were large (__Supplementary Figure S14B__, __Supplementary Tables S11__).
Collectively, the high co-linearity of the most important explanatory variables points suggests their redundancy. 
Tackling this redundancy e.g. by regularization may control the over-fitting and improve accuracy of predictions of DLCO and reduced DLCO by machine learning algorithms.

### 2025-03-06 paper/markdown/supplementary_material.Rmd
Variable redundancy
Inclusion of variable redundancy analysis data in Results of manuscript. modification of Supplementary Figure S14

## Co-linearity of the most influential explanatory variables

The most influential explanatory factors for predictions of DLCO (algorithms: random forest, GBM, neuronal network, and SVM radial) and DLCO < 80% of reference (algorithms: random forest, GBM, SVM radial) were identified as top 15 variables with the largest mean absolute SHAP importance values for each of the modeling algorithms. 
Associations between these highly influential variables were assessed with a correlation graph [@Csardi2006]. 
For construction of the correlation graph, ordered categorical variables (i.e. factors in R) were converted to integers. 
The graph edges were defined by pairwise correlations between the variables with Kendall's $\tau \geq 0.3$. 
The edge weights corresponded to values of the $\tau$ correlation coefficients. 
The graph was constructed with `as_iGraph()` function; isolated vertices, i.e. variables without any correlations with $\tau \geq$ 0.3 were removed from the graph with `prune_degree()` (both functions from [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra)). 
The graph was visualized with Fruchterman-Reingold algorithm (method `plot()`, package [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra)).

@@

flextable::flextable(suppl_tabs$lft_correlation_uni) %>% 
  set_widths(c(2.5, 2.5, 1.8, 4, 2.7)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: CT severity score determined by a human radiologist, sum for al lobes; opacity and high opacity, AI: opacity and high opacity of the lungs determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Blocked bootstrap Spearman correlation test; \u03C1 correlation coefficients with 95% confidence intervals. P values were corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

@@

_The most influential variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide by machine learning models were identified as features with the largest mean absolute Shapley additive explanations (SHAP) as presented in Figure 3 and Supplementary Figure S\@ref(fig:fig-varimp-dlco).__

@@

_(B) Correlations between computed tomography (CT) readouts of lung damage, the human-determined CT severity score and software-determined lung opacity and high opacity, were assessed by blocked bootstrap Kendall's correlation test. Ranks of the CT readouts are presented in scatter plots. Each point represents a single observation, blue straight lines are visualizations of linear trends. Correlation coefficients $\tau$ with bootstrap confidence intervals, p values, and numbers of observations are displayed in the plot captions._

_(C) Comparison of CTSS and lung opacity between observations with and without ground glass opacity (GGO) and reticulations. Statistical significance was determined by blocked bootstrap test with r effect size statistic. Median values with interquartile ranges are depicted as boxes, whiskers span over 150% of the interquartile ranges, single observations are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of observations are indicated in the X axes._

>>>>

## Co-linearity of explanatory variables

The most influential explanatory factors for predictions of DLCO (algorithms: random forest, GBM, neuronal network, and SVM radial) and DLCO < 80% of reference (algorithms: random forest, GBM, SVM radial) were identified as top 15 variables with the largest mean absolute SHAP importance values for each of the modeling algorithms. 
Associations between these highly influential variables were assessed with a soft-threshold weighted graph of correlations, a method widely used in omics and system biology [@Csardi2006; @Zhang2005]. 
For construction of the correlation graph, ordered categorical variables (i.e. factors in R) were converted to integers. 
The graph edges were defined by pairwise correlations between the variables with Kendall's $\tau \geq 0.3$. 
The edge weights corresponded to values of the $\tau$ correlation coefficients. 
The graph was constructed with `as_iGraph()` function; isolated vertices, i.e. variables without any correlations with $\tau \geq$ 0.3 were removed from the graph with `prune_degree()` (both functions from [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra)). 
The graph was visualized with Fruchterman-Reingold algorithm (method `plot()`, package [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra)).

@@

flextable::flextable(suppl_tabs$lft_correlation_uni) %>% 
  set_widths(c(3.5, 3.5, 1.8, 5, 2.7)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: CT severity score determined by a human radiologist, sum for al lobes; opacity and high opacity, AI: opacity and high opacity of the lungs determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Blocked bootstrap Spearman correlation test; \u03C1 correlation coefficients with 95% confidence intervals. P values were corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

@@

_The most influential variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide by machine learning models were identified as features with the largest mean absolute Shapley additive explanations (SHAP) as presented in Figure 3 and Supplementary Figure S\@ref(fig:fig-varimp-dlco)._

@@

_(B) Comparison of CTSS and lung opacity between observations with and without ground glass opacity (GGO) and reticulations. Statistical significance was determined by blocked bootstrap test with r effect size statistic. Median values with interquartile ranges are depicted as boxes, whiskers span over 150% of the interquartile ranges, single observations are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of observations are indicated in the X axes._
